A Plasma Torus Around a Young Low-Mass Star

### 2025-04-23 paper/ms.tex
update

The other analog for the CPVs are the $\sigma$~Ori~E variables, a rare
subset of B stars with radiatively-driven winds which accumulate into
warped plasma tori \citep{Townsend2005,Townsend2008}.  These tori tend
to have dense antipodal accumulations of plasma sculpted by
tilted-dipole magnetic fields, and the transits of these clumps produce
broadband photometric variability through bound-free scattering
\citep{Townsend2005} and Thomson scattering \citep{Berry2022}.  For
$\sigma$~Ori~E and almost all of its analogs, the result is simple
light curves that resemble those of eclipsing binaries, and time-dynamic
H$\alpha$ spectra similar to those in Figure~\ref{fig:spec}.  The two
known exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \citep{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong contributions
from higher-order magnetic moments \citep{Kochukhov2011,Shultz2018}.
The main implication is that the photometric complexity of CPVs may be a
direct consequence of magnetic fields with highly multipolar
contributions.

@@

The role of the star's magnetic field could be better understood through
new observations, and new theory.  From the modeling perspective,
there is an urgent need for rigid-field magnetohydrodynamic models to
go beyond previous work \citep{Townsend2005,Townsend2008,Krticka2022}
and to explore what field topologies might explain the observed light
curve morphologies.  In particular, dynamo simulations of
fully-convective M dwarfs have suggested that global-scale mean fields
might be confined to a single hemisphere \citep{Brown2020}; such fields
would yield accumulation surfaces quite different from those that have
previously been explored.  Observationally, spectropolarimetry has the
potential to assess both the field strength and topology.  An
independent constraint on magnetic field topology may also come from
radio observations, which have shown \citep{Kaur2024} that CPVs emit
variable radio signals, including persistent and polarized components.
Detecting radio emission produced by an electron cyclotron maser
instability (e.g.~\citealt{Callingham2021}) in particular would provide
a measurement of the field strength at the site of the emitting region,
further clarifying the magnetosphere's structure.

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs
\citep{Dressing2015}.  However, a few percent of young M dwarfs show
the CPV phenomenon \citep{Rebull2020}, and our data show that some of
the complex photometric features occur when clumps of circumstellar
material transit the star.  The suggested geometric correction implies
that an appreciable minority ($\gtrsim$10\%) of young M dwarfs -- the
rapidly rotating ones with centrifugal magnetospheres -- host
circumstellar environments similar to the CPVs.  Future studies that
combine spectroscopic, polarimetric, and multi-wavelength
observations, along with magnetohydrodynamic modeling, will be key to
understanding the complex magnetospheric environments of these young
stars.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.48\textwidth]{figures/sf4.pdf}
  \caption{
    Spectral energy distribution of broadband photometric magnitudes
    (filled markers) plotted over the best-fit BT-Settl stellar
    atmosphere model \citep{Allard2012} and the associated photometric
    predictions (empty markers).  This plot was made from an
    adaptation of \texttt{astroARIADNE} \citep{Vines2022}.  The
    photometry extends from the Gaia DR3 blue passband to WISE W3;
    the W4 passband (22 $\mu$m) did not yield a confident detection.
    This fit yields the star's temperature and size.  The lack of
    excess infrared flux relative to the photospheric model sets an
    upper limit on emission from circumstellar dust.
    }
  \label{fig:sed}
\end{figure}

@@

\begin{figure*}[!tp]
  \centering
  \includegraphics[width=0.55\textwidth]{figures/sf6.pdf}
  \caption{Time-variable fit to H$\alpha$ line profiles.  {\it Left
  column:} Raw spectrum at each epoch $f_{\lambda}$ minus the
  time-averaged spectrum $f_{\langle t \rangle}$ (as in
  Figure~\ref{fig:spec}e).  Underplotted sinusoids are not fits; they
  are meant to guide the eye.  {\it Middle columns:} Model of emission
  from the inner clump (sum of two gaussians) and the outer clump
  (single gaussian), plotted over the data.  {\it Right column:}
  Residual of the left column after subtracting the sum of the two
  middle columns, leaving variability in the line core.
  Appendix~\ref{subsec:halpha} discusses the use of this model.  }
  \label{fig:halphamodel}
\end{figure*}

@@

\begin{figure*}[!t]
  \centering
  \includegraphics[width=\textwidth]{figures/sf5.pdf}
  \caption{Time evolution of select regions in the Keck/HIRES spectra
  from 17 February 2024.  We made this plot by applying a windowed
  outlier rejection to remove cosmic rays and then smoothing each
  spectrum with a Gaussian filter.  The horizontal axis shows the
  velocity relative to each line's rest wavelength, normalized by the
  stellar equatorial velocity.  H$\alpha$ is the only Balmer line to
  show periodic variability of the form seen in Figure~\ref{fig:spec}.
  The He 5876\,\AA\ line shows a time-dynamic blueshift that differs
  from the Balmer lines.  Li 6708\,\AA\ shows no obvious absorption. }
  \label{fig:hirescuts}
\end{figure*}

@@

Figure~\ref{fig:hirescuts} provides a novel view on the
blue excess that appeared at $\phi$$\approx$0.2 in H$\alpha$,
H$\gamma$, H$\delta$, and He 5876\,\AA.  At the same epoch, H$\alpha$
additionally shows a red clump of emission, and H$\gamma$ and
H$\delta$ are also broadened on the red wing.  The rise of this
emission in $<$15\,minutes might suggest a more sudden flow, rather
than a stable, periodic component.  For instance, a stellar flare
might be connected to such a sudden rise.  However, this idea seems
incompatible with the sinusoidal emission seen later from
$\phi$$\approx$0.5-1.0, and with the fact that flares typically excite
iron lines in the blue HIRES orders, which are not observed.  More
time-series spectroscopy would be needed to clarify this type of
variability, in particular whether it is periodic or stochastic.

>>>>

The other analog for the CPVs are the $\sigma$~Ori~E variables, a rare
subset of B stars with radiatively-driven winds which accumulate into
warped plasma tori \citep{Townsend2005,Townsend2008}.  These tori tend
to have dense antipodal accumulations of plasma sculpted by
tilted-dipole magnetic fields, and the transits of these clumps produce
broadband photometric variability through bound-free scattering
\citep{Townsend2005} and Thomson scattering \citep{Berry2022}.  For
$\sigma$~Ori~E and almost all of its analogs, the result is simple
light curves that resemble those of eclipsing binaries, and time-dynamic
H$\alpha$ spectra similar to those in Figure~\ref{fig:spec}.  The two
known exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \citep{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong contributions
from higher-order magnetic moments \citep{Kochukhov2011,Shultz2018}.
This suggests that the photometric complexity of CPVs may be related to
magnetic fields with highly multipolar contributions.

@@

The role of the star's magnetic field could be better understood through
new observations, and new theory.  From the modeling perspective,
there is a need for rigid-field magnetohydrodynamic models to
go beyond previous work \citep{Townsend2005,Townsend2008,Krticka2022}
and to explore what field topologies explain the observed light
curve shapes.  In particular, dynamo simulations of
fully-convective M dwarfs have suggested that global-scale mean fields
might be confined to a single hemisphere \citep{Brown2020}; such fields
would yield accumulation surfaces quite different from those that have
previously been explored.  Observationally, spectropolarimetry has the
potential to assess both the field strength and topology.  An
independent constraint on magnetic field topology may also come from
radio observations, which have shown \citep{Kaur2024} that CPVs emit
variable radio signals, including persistent and polarized components.
Detecting radio emission produced by an electron cyclotron maser
instability (e.g.~\citealt{Callingham2021}) in particular would provide
a measurement of the field strength at the site of the emitting region,
which could help clarify the magnetosphere's structure.

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs
\citep{Dressing2015}.  However, a few percent of young M dwarfs show
the CPV phenomenon \citep{Rebull2020}, and our data show that some of
the complex photometric features occur when clumps of circumstellar
material transit the star.  The suggested geometric correction implies
that an appreciable minority ($\gtrsim$10\%) of young M dwarfs -- the
rapidly rotating ones with centrifugal magnetospheres -- host
circumstellar environments similar to the CPVs.  Future studies that
combine spectroscopic, polarimetric, and multi-wavelength
observations, along with magnetohydrodynamic modeling, will be key to
understanding the complex environments of these young
stars.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.48\textwidth]{figures/sf4.pdf}
  \caption{
    {\bf Spectral energy distribution} of broadband photometric magnitudes
    (filled markers) plotted over the best-fit BT-Settl stellar
    atmosphere model \citep{Allard2012} and the associated photometric
    predictions (empty markers).  This plot was made from an
    adaptation of \texttt{astroARIADNE} \citep{Vines2022}.  The
    photometry extends from the Gaia DR3 blue passband to WISE W3;
    the W4 passband (22 $\mu$m) did not yield a confident detection.
    This fit yields the star's temperature and size.  The lack of
    excess infrared flux relative to the photospheric model sets an
    upper limit on emission from circumstellar dust.
    }
  \label{fig:sed}
\end{figure}

@@

\begin{figure*}[!tp]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/sf6.pdf}
  \caption{{\bf Time-variable fit to H$\alpha$ line profiles.}  {\it Left
  column:} Raw spectrum at each epoch $f_{\lambda}$ minus the
  time-averaged spectrum $f_{\langle t \rangle}$ (as in
  Figure~\ref{fig:spec}e).  Underplotted sinusoids are not fits; they
  are meant to guide the eye.  {\it Middle columns:} Model of emission
  from the inner clump (sum of two gaussians) and the outer clump
  (single gaussian), plotted over the data.  {\it Right column:}
  Residual of the left column after subtracting the sum of the two
  middle columns, leaving variability in the line core.
  Appendix~\ref{subsec:halpha} discusses the use of this model.  }
  \label{fig:halphamodel}
\end{figure*}

@@

\begin{figure*}[!t]
  \centering
  \includegraphics[width=\textwidth]{figures/sf5.pdf}
  \vspace{-0.25cm}
  \caption{{\bf Time evolution of the Keck/HIRES spectra
  from 17 February 2024.}  We made this plot by applying a windowed
  outlier rejection to remove cosmic rays and then smoothing each
  spectrum with a Gaussian filter.  The horizontal axis shows the
  velocity relative to each line's rest wavelength, normalized by the
  stellar equatorial velocity.  H$\alpha$ is the only Balmer line to
  show periodic variability similar to Figure~\ref{fig:spec}.
	The He 5876\,\AA\ line shows a time-dynamic blueshift.  Li 6708\,\AA\ shows
	no obvious absorption. }
  \label{fig:hirescuts}
\end{figure*}

@@

Figure~\ref{fig:hirescuts} provides a novel view on the
blue excess that appeared at $\phi$$\approx$0.2 in H$\alpha$,
H$\gamma$, H$\delta$, and He 5876\,\AA.  At the same epoch, H$\alpha$
additionally shows a red clump of emission, and H$\gamma$ and
H$\delta$ are also broadened on the red wing.  The rise of this
emission in $<$15\,minutes might suggest a more sudden flow, rather
than a stable, periodic component.  For instance, a stellar flare
might be connected to such a sudden rise.  However, this idea seems
incompatible with the sinusoidal emission seen from
$\phi$$\approx$0.3-1.0, and with the fact that flares typically excite
iron lines in the blue HIRES orders, which are not observed.  More
time-series spectroscopy would be needed to clarify this type of
variability, in particular whether it is periodic or stochastic.
The gap in the TESS photometry (Figure~\ref{fig:fulllc}) additionally
clouds the interpretability of this event.
